---
title: "Selection constrains high rates of satellite DNA mutation in Daphnia pulex"
author: "Jullien Flynn"
date: '2017-01-12'
output:
  html_document:
    toc: yes
  pdf_document:
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=60,tidy=TRUE))
```
**Daphnia kmer Analysis for Flynn et al. 2017 manuscript**  

GC correction bins used: [0, 0.25, 0.30, 0.35, 0.40, 0.45, 0.50, 0.55, 0.60, 0.75, 1.0]  

Load necessary packages:
``` {r, load packages, include=FALSE}
## load packages
rm(list = ls()) # clear environment
library(devtools)
library(ggbiplot)
library(raster)
library(corrplot)
source("http://bioconductor.org/biocLite.R")
require(Biostrings)
library(adegenet)
library(ape)
library(Rcmdr)

```
# Import data
This is already normalized by GC and depth by python scripts available in the GC_correction folder on github.  

First, perform pre-processing on the data for later steps.  
```{r, load data}
daphniakmers_normalized_w_techrep <- read.csv("~/Documents/FirstRotation-Andy/daphniakmers_corrected_bins3_Jan7.csv")

file_names <- as.vector(daphniakmers_normalized_w_techrep$X)
file_names <- sapply(strsplit(file_names, "\\."), '[', 1) 
rownames(daphniakmers_normalized_w_techrep) <- file_names
# now get rid of the column that has the file name
daphniakmers_normalized_w_techrep[,1] <- NULL
# the columns (kmers) by their mean abundance
means <- colMeans(daphniakmers_normalized_w_techrep)
index.orders <- order(means, decreasing=T)
daphniakmers_normalized_w_techrep <- daphniakmers_normalized_w_techrep[,order(means, decreasing=T)]
# remove the total bp column and the poly-N columns
daphniakmers_normalized_w_techrep$total_bp <- NULL
daphniakmers_normalized_w_techrep$N.N <- NULL

## need to convert to matrix
daphniakmers_normalized_w_techrep <- as.matrix(daphniakmers_normalized_w_techrep)

# this is everything, with the technical replicates.

# remove the technical replicates from this analysis for now, will come back to them in the quality control analysis file.
# technical replicates are called C001 and C035 (from C01 and C35)
techreps <- c("C001", "C035")

daphniakmers_normalized <- daphniakmers_normalized_w_techrep[!rownames(daphniakmers_normalized_w_techrep) %in% techreps, ]
```
## Set the identities of the different samples
```{r}
MA1_names <- c("C01", "C02", "C03", "C06", "C07", "C08", "C12", "C13", "C14", "C16", "C17", "C18", "C20", "C21", "C24", "C25", "C34", "C35", "C36", "C37", "C38", "C39", "C40", "C44")
MA1_indexes <- match(MA1_names, rownames(daphniakmers_normalized))
MA2_names <- c("C23", "C27", "C43", "C49")
MA2_indexes <- match(MA2_names, rownames(daphniakmers_normalized))
Pop_names <- c("CC3", "CC4", "CC6", "CC7", "CC8", "CC9")
Pop_indexes <- match(Pop_names, rownames(daphniakmers_normalized))

```
# Description of kmers in Daphnia

## Get common kmers  
Common kmers have a mean abundance of at least 100 copies across all lines. These are the kmers that we will use for most analyses and comparisons.   
Perform a few summaries on the common kmers:  

```{r, common kmers}
# get the common kmers
common.kmer.indexes <- c()
for (i in 1:ncol(daphniakmers_normalized)) {
  if (mean(daphniakmers_normalized[,i])>=100) {
    common.kmer.indexes <- c(common.kmer.indexes, i)
  }
}
names(common.kmer.indexes) <- colnames(daphniakmers_normalized)[common.kmer.indexes]
common.kmer.names <- colnames(daphniakmers_normalized)[common.kmer.indexes]

# number of kmers with at least 100 mean normalized copies.
length(common.kmer.indexes) 

# function that calculates the length of a kmer
Calc_kmer_length <- function (kmer_string) {
  return (nchar(kmer_string))
}
# calculate the lengths of the common kmers
common.kmer.lengths <- sapply(colnames(daphniakmers_normalized)[common.kmer.indexes], Calc_kmer_length) 
names(common.kmer.lengths) <- colnames(daphniakmers_normalized)[common.kmer.indexes]

# function that calculates the GC content
Calc_GC_content <- function (kmer_string) {
  string <- BString(kmer_string)
  gc.content <- c(letterFrequency(string, letters=c("CG"), as.prob=F)/(length(string)))
  return(gc.content)
}
# calculate the GC content of the common kmers
common.kmer.GC <- sapply(colnames(daphniakmers_normalized)[common.kmer.indexes], Calc_GC_content)
names(common.kmer.GC) <- colnames(daphniakmers_normalized)[common.kmer.indexes]
# visualize the GC content histogram
hist(common.kmer.GC, breaks=12, main="", xlab="GC content")

# which kmers have at least 1000 copies?
kb.kmer.indexes <- c()
for (i in 1:ncol(daphniakmers_normalized)) {
  if (mean(daphniakmers_normalized[,i])>=1000) {
    kb.kmer.indexes <- c(kb.kmer.indexes, i)
  }
}
length(kb.kmer.indexes)
common.kmer.names[kb.kmer.indexes] # these are the kmers that have at least 1000 copies
```
##Get the kmers that have at least 2 copies  
These will be called the "legit kmers" because they are present in all lines at at least 2 copies after normalization (in reality, in order to be detected by k_Seek, kmers must be present in a tandem array spanning at least 50 bp of the read).  

``` {r}
# all the legit kmers
legit.kmer.indexes <- c()
for (i in 1:ncol(daphniakmers_normalized)) { ## go through each column of the data
  p <- daphniakmers_normalized[,i] >= 2 ## at least 2 copies
  if (sum(p)==nrow(daphniakmers_normalized)) { # if all samples have at least 2 copies
    legit.kmer.indexes <- c(legit.kmer.indexes, i)
  }
}
length(legit.kmer.indexes) # number of legit kmers
legit.kmer.names <- colnames(daphniakmers_normalized)[legit.kmer.indexes]

```
## Mean counts, length summary, absolute satDNA content
```{r, kmer summaries}

# get the means of the common kmers
common.kmer.means <- colMeans(daphniakmers_normalized[,common.kmer.indexes])

# summary of the lengths
summary(as.factor(common.kmer.lengths))

# Absolute kmer content
#use the "legit kmers" to add up the entire kmer content across lines
kmers.abs <- matrix(nrow=nrow(daphniakmers_normalized), ncol=max(legit.kmer.indexes))
for (i in legit.kmer.indexes) {
  for (j in 1:nrow(daphniakmers_normalized)) {
    kmers.abs[j,i] <- daphniakmers_normalized[j,i] * (nchar(colnames(daphniakmers_normalized)[i], type="chars")) 
  }
}
kmers.abs <- kmers.abs[,colSums(is.na(kmers.abs)) != nrow(kmers.abs)] # get rid of columns that are all NAs

total.kmers.abs <- c()
for (i in 1:nrow(kmers.abs)){
  total.kmers.abs[i] <- sum(kmers.abs[i,])/10^6
}
names(total.kmers.abs) <- rownames(daphniakmers_normalized)

# These absolute kmer contents are in Megabases. 
total.kmers.abs
min(total.kmers.abs)
max(total.kmers.abs)
median(total.kmers.abs)
# note: will return to this later when comparing MA lines with population

# Look further into the length of the kmers. There were no 15-mers in the common kmers, how many are there in the lower abundance, but still present, "legit kmers"?
legit.kmer.lengths <- sapply(legit.kmer.names, Calc_kmer_length)
length(which(legit.kmer.lengths==15))
length(which(legit.kmer.lengths==5))
length(which(legit.kmer.lengths==10))
length(which(legit.kmer.lengths==20))
length(which(legit.kmer.lengths==12))
hist(legit.kmer.lengths, breaks=20)

```


## Output the kmers into tableS1
``` {r}
tableS1 <- data.frame(c(1:length(legit.kmer.names)), legit.kmer.names, colMeans(daphniakmers_normalized[,legit.kmer.indexes]))
colnames(tableS1) <- c("kmer index", "kmer sequence", "mean abundance")
rownames(tableS1) <- NULL

#write.table(tableS1, file="" , quote=F, col.names=T, row.names=F, sep="\t")
```
## Figure 1
```{r}
# get the standard deviation for error bars.
common.kmer.sd <- apply(daphniakmers_normalized[,common.kmer.indexes], 2, sd)
# make a table
kmer.table <- data.frame(common.kmer.names, common.kmer.means, common.kmer.lengths, common.kmer.sd)
# order the kmers by kmer length
ordered.kmer.table <- kmer.table[order(kmer.table$common.kmer.lengths),]
# make a barplot that will be Figure 1 of the manuscript
par(mar=c(8,4,1,1))
fig1 <- barplot(ordered.kmer.table$common.kmer.means, log="y", ylim=c(100,60000), names.arg=ordered.kmer.table$common.kmer.names, las=2, cex.names=0.4, cex.axis=0.8, ylab="normalized abundance", col="grey" )
arrows(x0=fig1, y0=(ordered.kmer.table$common.kmer.means-ordered.kmer.table$common.kmer.sd), x1=fig1, y1=ordered.kmer.table$common.kmer.means + ordered.kmer.table$common.kmer.sd, angle=90, code=3, length=0)

```
  
## Comparison to D. melanogaster
Compare to what was found in Drosophila melanogaster
Use kmers found in at least 100 copies in D. mel female (don't want to have Y-specific kmers included).  File of D. mel kmers was obtained from unpublished data from KH-C Wei, SS Lower, JJ Sless, and DA Barbash. 

For python scripts used for comparisons, see MotifSearch folder on github.  


```{r, dmel comparison}
# import data
dmelkmers <- read.delim("~/Documents/FirstRotation-Andy/dmel.normalizedkmers.txt") # already normalized, already only from D. mel female.
# get the ones with at least 100 bp
dmel.common.indexes <- c()
for (i in 1:ncol(dmelkmers)) {
  if (dmelkmers[1,i]>=100) {
   dmel.common.indexes <- c(dmel.common.indexes, i)
  }
}

dmel.common.kmers <- names(dmelkmers)[dmel.common.indexes]
dmel.common.kmers <- dmel.common.kmers[2:length(dmel.common.kmers)]
#write(dmel.common.kmers, file="")


```
# Mutation rates in the MA lines  
Use the mean abundance of each kmer of the population as the ancestral state of the population.  

```{r, mutation rates}
# define the MA and non-MA population indexes
MA.indexes <- c(1:28) 
Pop.indexes <- c(29:34)
# make a new matrix that calculates the deviation from the mean abundance of the population
kmer.deviation <- matrix(nrow=nrow(daphniakmers_normalized), ncol=length(common.kmer.indexes))
for (i in common.kmer.indexes) {
  for (j in 1:nrow(daphniakmers_normalized)){
    kmer.deviation[j,i] <- daphniakmers_normalized[j,i] - mean(daphniakmers_normalized[Pop.indexes,i])
  }
}

colnames(kmer.deviation) <- common.kmer.names
rownames(kmer.deviation) <- rownames(daphniakmers_normalized)

# import file that contains the generations information
generations_file <- read.table(file="~/Documents/FirstRotation-Andy/generations.txt", header=T)

generations <- as.vector(generations_file[,2])
names(generations) <- generations_file[,1]
```
## Kmer abundance means similar in MA lines as Population

``` {r}
# calculate the means of each kmer for the MA lines and the population isolates
MA.kmer.means <- colMeans(daphniakmers_normalized[MA.indexes, common.kmer.indexes])
Pop.kmer.means <- colMeans(daphniakmers_normalized[Pop.indexes, common.kmer.indexes])

hist(MA.kmer.means, breaks=8, xlab="Mean abundance in MA lines")
hist(Pop.kmer.means, breaks=8, xlab="Mean abundance in Population")


# quantify the amount of variation in kmer mean abundance between MA lines and pop 
MA.kmer.sd <- apply(daphniakmers_normalized[MA.indexes, common.kmer.indexes], 2, sd)

sum(Pop.kmer.means > (MA.kmer.means + (2*MA.kmer.sd)))
sum(Pop.kmer.means < (MA.kmer.means - (2*MA.kmer.sd)))

sum(Pop.kmer.means > (MA.kmer.means + (1*MA.kmer.sd)))
sum(Pop.kmer.means < (MA.kmer.means - (1*MA.kmer.sd)))
# 27/39 within 1 SD
# all within 2 SD

```

## Box plot
``` {r, box plot}
#get in format for box plot, each MA line an observation
# not including the telomere repeat here, which is index 2.
mutation.rate.data <- c()
mutation.kmer.indexes <- common.kmer.indexes[c(1,3:length(common.kmer.names))]
for (j in mutation.kmer.indexes) {
  for (i in MA.indexes) {
    mutation.rate.data <- c(mutation.rate.data, (kmer.deviation[i,j]/mean(daphniakmers_normalized[Pop.indexes,j])/generations[i]))
  }
}

mutation.df <- data.frame(mutation.rate.data, common.kmer.names[mutation.kmer.indexes])
mutation.df$common.kmer.names.mutation.kmer.indexes. <- factor (mutation.df$common.kmer.names.mutation.kmer.indexes., levels = common.kmer.names[mutation.kmer.indexes])

par(mar=c(8,4,1,1))
boxplot (mutation.df[,1] ~ mutation.df[,2], data=mutation.df, las=2, cex.axis=0.5, ylab="Copies changed/gen/copy")
abline(h=0, col="red", lwd=2)

```
  
It seems more expansions than contractions  
Look across all kmers to see if this is indeed the case - might be biased to expansions in the common kmers.  
Actually, this pattern of expansions in the MA lines compared to the population is across all kmers.  
  
## Expansions more common than contractions
```{r, expansions and contractions}
# deviation for all kmers
kmer.deviation.all <- matrix(nrow=nrow(daphniakmers_normalized), ncol=ncol(daphniakmers_normalized))
for (i in 1:ncol(daphniakmers_normalized)) {
  for (j in 1:nrow(daphniakmers_normalized)){
    kmer.deviation.all[j,i] <- daphniakmers_normalized[j,i] - mean(daphniakmers_normalized[Pop.indexes,i])
  }
}

colnames(kmer.deviation.all) <- colnames(daphniakmers_normalized)
rownames(kmer.deviation.all) <- rownames(daphniakmers_normalized)

# calculate the mutation rate across all legit kmers.
mutation.rate.all <- matrix(ncol=ncol(kmer.deviation.all), nrow=length(MA.indexes))
for (j in legit.kmer.indexes) {
  for (i in MA.indexes) {
    mutation.rate.all[i,j] <- kmer.deviation.all[i,j]/mean(daphniakmers_normalized[Pop.indexes,j])/generations[i]
  }
}

mutation.rate.all <- mutation.rate.all[,colSums(is.na(mutation.rate.all)) != nrow(mutation.rate.all)]
colnames(mutation.rate.all) <- legit.kmer.names
rownames(mutation.rate.all) <- rownames(daphniakmers_normalized)[MA.indexes]

## now for each kmer, calcalate the average, the number of lines that expanded, and the number that contracted

mutation.rate.all.summary <- matrix(ncol=length(legit.kmer.indexes), nrow=3)
for (i in 1:ncol(mutation.rate.all)) {
  nexp <- 0
  ncontr <- 0
  avg <- mean(mutation.rate.all[,i])
  for (j in 1:nrow(mutation.rate.all)) {
    if (mutation.rate.all[j,i] > 0) {
      nexp <- nexp + 1
    } else {
      ncontr <- ncontr + 1
    }
  }
  mutation.rate.all.summary[1, i] <- avg
  mutation.rate.all.summary[2, i] <- nexp
  mutation.rate.all.summary[3, i] <- ncontr
}
colnames(mutation.rate.all.summary) <- legit.kmer.names
rownames(mutation.rate.all.summary) <- c("avg", "Nexp", "Ncontr")
mutation.rate.all.summary[,1:10]

plot(mutation.rate.all.summary[1,], xlab="Kmer index", ylab="copies changed/gen/copy")
abline(h=0, col="red")

length(which(mutation.rate.all.summary[2,] > mutation.rate.all.summary[3,])) # 109 of 162 have more lines with expansions than contractions
length(which(mutation.rate.all.summary[2,] < mutation.rate.all.summary[3,])) # 46 have more MA lines with contractions than MA lines with expansions
length(which(mutation.rate.all.summary[1,] > 0)) # 124 of 162 have overall average expansion

```
## Calculate overall mutation rates per kmer  
Take the absolute value of the deviation: this is the sum of expansions/contractions.  
``` {r}
kmer.deviation.abs <- abs(kmer.deviation)
mean_generations <- mean(generations)
# sum all the numbers of copy changes and average
mutation.rate.abs <- colMeans(kmer.deviation.abs[MA.indexes, mutation.kmer.indexes])/mean_generations
sort(mutation.rate.abs, decreasing=T)
median(mutation.rate.abs)
mean(mutation.rate.abs)

# see if correlation with kmer abundance
plot(y=mutation.rate.abs, x=common.kmer.means[mutation.kmer.indexes], pch=19,  xlab="Mean abundance", ylab="Copies/generation", col="tan1", cex=1.4)
abline(lm(formula = mutation.rate.abs ~ common.kmer.means[mutation.kmer.indexes]), col="red", lwd=2)

# seems to be linear relationship, check statistical significance
summary(lm(formula = mutation.rate.abs ~ common.kmer.means[mutation.kmer.indexes]))

# no relationship with kmer length
plot(y=mutation.rate.abs, x=common.kmer.lengths[mutation.kmer.indexes], log="y", xlab="kmer length", ylab="mutation rate", pch=19)


# no relationship with GC content
plot(y=mutation.rate.abs, x=common.kmer.GC[mutation.kmer.indexes], log="y", xlab="GC content", ylab="mutation rate", pch=19)

# What about the telomere?
telomere.change.rate <- mean(kmer.deviation.abs[MA.indexes, 2])/mean_generations
mean(daphniakmers_normalized[MA.indexes,2])
mean(daphniakmers_normalized[Pop.indexes,2])
```
## Normalize the mutation rate by the kmer abundance  
Because more abundant kmers might have a higher mutation rate.  
Normalize so rates are expressed as changes/generations/copy.
```{r}
initial_means <- c()
for (j in common.kmer.indexes) {
  initial_means <- c(initial_means, mean(daphniakmers_normalized[Pop.indexes,j]) )
}
names(initial_means) <- colnames(kmer.deviation) 

mutation.rate.abs.norm <- colMeans(kmer.deviation.abs[MA.indexes,mutation.kmer.indexes])/(mean_generations*initial_means[mutation.kmer.indexes])
plot(mutation.rate.abs.norm)
# come back to plot this when comparing to the Population isolates
mean(mutation.rate.abs.norm)
median(mutation.rate.abs.norm)
which(mutation.rate.abs.norm == max(mutation.rate.abs.norm))
min(mutation.rate.abs.norm)
max(mutation.rate.abs.norm)

# come back to plot this when comparing to the Population isolates

```

## Kmers generated de novo
Search for kmers that are present in 1-2 MA lines but no others.  
First make a new matrix with all kmers that are present in at least 1 normalized copy in at least 1 MA line. 

```{r}
daphniakmers.all <- as.matrix(daphniakmers_normalized[,common.kmer.indexes]) # start with small matrix and add to it
for (i in (max(common.kmer.indexes)+1):ncol(daphniakmers_normalized)){ ## go through all columns in data
  counter <- 0
  for (j in 1:nrow(daphniakmers_normalized)) {
    if (daphniakmers_normalized[j,i] > 2 ) {
      counter <- counter + 1
    }
  }
  if (counter > 0) { ## if there is at least one sample with 2 copies of this kmer, add it to the new matrix
    daphniakmers.all <- cbind(daphniakmers.all[, , drop=FALSE], daphniakmers_normalized[,i, drop=FALSE])
  }
}
ncol(daphniakmers.all) # all kmers that have at least one line with 3 copies normalized.
```
### Get the new kmers
```{r}

get_new_kmers <- function (nlines) {
  kmers.new <- matrix(nrow=nrow(daphniakmers.all)) # make new empty matrix
  for (i in 1:ncol(daphniakmers.all)) { ## go through each column of the data
    uniq <- daphniakmers.all[,i] > 0 # uniq will be true and false 
    if (sum(uniq)==nlines) { # if only there are nlines true elements, add to matrix
      kmers.new <- cbind(kmers.new[, , drop=FALSE], daphniakmers.all[,i, drop=FALSE])
    }
  }
  kmers.new <- kmers.new[,2:ncol(kmers.new)] # remove first column, which is all NA
  return(kmers.new)
}

new_kmers_unique <- get_new_kmers(1)
colnames(new_kmers_unique)
new_kmers_pairs <- get_new_kmers(2)
colnames(new_kmers_pairs)

```
## Get the mean abundance of the similar kmer
``` {r}
# these are the kmers that are already at high abundance and seem to be similar to the new kmers
similar_kmers <- c("AATGG", "AAAAG", "AACAG", "AACCT", "AAAG", "AAGG", "AAAG")

similar_kmer_abundances <- c()
for (i in 1:length(similar_kmers)) {
  index <- which(legit.kmer.names == similar_kmers[i])
  abun <- mean(daphniakmers_normalized[,index])
  similar_kmer_abundances <- c(similar_kmer_abundances, abun)
}
names(similar_kmer_abundances) <- similar_kmers

similar_kmer_abundances

which(colnames(daphniakmers_normalized)=="ACAACCCACTAGGGTTT")
```
### Quality control - look for kmers unique to library prep  
No kmers are unique to one library prep.

``` {r}
get_lib2_uniq_kmers <- function (indexes_to_include, indexes_to_exclude) {
  kmers.uniq <- matrix(nrow=34)
  for (i in 1:ncol(daphniakmers.all)) { ## go through each column of the data
    uniq <- daphniakmers.all[indexes_to_include,i] > 0 # gives vector of true and falses
    absent <- daphniakmers.all[indexes_to_exclude, i] == 0
    if (sum(uniq)==6 & sum(absent)==28) { 
      kmers.uniq <- cbind(kmers.uniq[, , drop=FALSE], daphniakmers.all[,i, drop=FALSE])
    }
  }
  if (ncol(kmers.uniq)<2) { 
    return ("none")  
  } else {
    return(kmers.uniq)
    }
}
lib1.indexes <- MA1_indexes
lib2.indexes <- c(MA2_indexes, Pop_indexes)
get_lib2_uniq_kmers(lib2.indexes, lib1.indexes) 
get_lib2_uniq_kmers(lib1.indexes, lib2.indexes)
```
  
## Look for kmers LOST de novo  

```{r}
# function returns the kmers lost in n lines
get_kmers_lost <- function (nlines) {
  kmers.lost <- matrix(nrow=34)
  for (i in 1:ncol(daphniakmers.all)) { ## go through each column of the data
    l <- daphniakmers.all[,i] == 0  ## 0 copies
    p <- daphniakmers.all[,i] > 1 ## present
    if (sum(l)==nlines && sum(p)==(34-nlines)) { # if there is only nlines line with 0 copies, and all others are present
      kmers.lost <- cbind(kmers.lost[, , drop=FALSE], daphniakmers.all[,i, drop=FALSE])
    }
  }
  kmers.lost <- kmers.lost[,2:ncol(kmers.lost)] # first column is NA
}

kmers_lost_uniq <- get_kmers_lost(1)
ncol(kmers_lost_uniq)
kmers_lost_uniq
kmers_lost_pairs <- get_kmers_lost(2)
ncol(kmers_lost_pairs)
kmers_lost_pairs

# for lost kmers, calculate the mean in the other lines
Calc_lost_other_means <- function (table) {
  other_means <- c()
  for (i in 1:ncol(table)) {
  other_means <- c(other_means, mean(table[which(table[,i]>0),i]))
  }
  return(other_means)
}
Calc_lost_other_means(kmers_lost_uniq)
Calc_lost_other_means(kmers_lost_pairs)
```
  
# Compare MA lines with population

## Compare the total satDNA content

``` {r}
min(total.kmers.abs[MA.indexes])
min(total.kmers.abs[Pop.indexes])
max(total.kmers.abs[MA.indexes])
max(total.kmers.abs[Pop.indexes])
median(total.kmers.abs[MA.indexes])
median(total.kmers.abs[Pop.indexes])

max(total.kmers.abs[MA.indexes])/min(total.kmers.abs[MA.indexes]) 
# MA lines diverged by a factor of 1.66

max(total.kmers.abs[Pop.indexes])/min(total.kmers.abs[Pop.indexes])
# Population isolates diverged by only factor of 1.27

## make a box plot of this.
vals <- c(total.kmers.abs[MA.indexes], total.kmers.abs[Pop.indexes])
cat <- c( rep("MA lines", times=28), rep("Population", times=6))
kmers.abs.df <- data.frame(vals, cat) 
boxplot (kmers.abs.df[,1] ~ kmers.abs.df[,2], data=kmers.abs.df, cex.axis=0.8, ylab="Absolute kmer content (Mb)")

# not a significant difference based on the Levene's test 

groups <- as.factor(c(rep(1, times=28), rep(2, times=6)))
leveneTest(total.kmers.abs, groups, center=median)
```
### Subsample from the MA lines  
```{r}


subsample_totalcontent <- function (CI) {
  means <- c()
  variances <- c()
  for (j in 1:1000) {
    k <- sample(MA.indexes, size=6, replace=T)
    means <- c(means, mean(total.kmers.abs[k]))
    variances <- c(variances, var(total.kmers.abs[k]))
  }
  quantiles_means <- quantile(means, CI)
  quantiles_vars <- quantile(variances, CI)
  output <- c (quantiles_means, quantiles_vars)
  return(output)
}

subsample_totalcontent(0.05)
mean (total.kmers.abs[Pop_indexes]) # below the 5%
var (total.kmers.abs[Pop_indexes]) # not below the 5%

subsample_totalcontent(0.1)
mean (total.kmers.abs[Pop_indexes]) # below the 10%
var (total.kmers.abs[Pop_indexes]) # not below the 10%

```

## Tendancy to lose or gain  
Question: If an MA experiences expansions in one kmer, does this mean it has the tendancy to expand across all kmers, or does it mean it has trade-offs in other kmers such that there is a balance of expansions/contractions within the same genome?  

``` {r}
# calculate the total deviation, including the sign of deviation from the ancestral state
total.deviation <- c()
for (i in 1:nrow(kmer.deviation)) {
  total.deviation <- c(total.deviation, sum(kmer.deviation[i,]))
}
#Fig 4B
# histogram showing the distribution of cumulative gains/losses across MA lines and pop isolates.  
names(total.deviation) <- rownames(kmer.deviation)
hist(total.deviation[MA.indexes], col=rgb(1,0,0,0.5), main="", xlab="Total copy number change across all kmers", breaks=10, xlim=c(-80000, 80000))
hist(total.deviation[Pop.indexes], col="blue", add=T)
box()

# other way to visualize this data
par(mar=c(5,4,4,1))
plot(total.deviation[MA.indexes], ylim=c(-max(total.deviation), max(total.deviation)), pch=19, col="red", ylab="cumulative copy changes", xaxt="n", xlab="")
points(y=total.deviation[Pop.indexes], x=c(4,8,12,16,24,28), pch=19, col="blue")
abline(h=0)

# one MA line is standing out
gainer <- which(total.deviation==max(total.deviation))

# check how many kb it gained
kmer.deviation[gainer, common.kmer.indexes]

gainer.total.gains <- 0
for (i in common.kmer.indexes) {
  gainer.total.gains <- gainer.total.gains + (kmer.deviation[gainer, i])*nchar(colnames(kmer.deviation)[i])
}
gainer.total.gains
sum(kmer.deviation[gainer,] < 0) # how many kmers did it experience gains in
```
  
## Compare the Coefficent of Variation (CV)  

```{r}

MA.cv <- c()
for (i in common.kmer.indexes) {
  MA.cv <- c(MA.cv, cv(daphniakmers_normalized[MA.indexes,i]))
}
names(MA.cv) <- common.kmer.names

Pop.cv <- c()
for (i in common.kmer.indexes){
  Pop.cv <- c(Pop.cv, cv(daphniakmers_normalized[Pop.indexes,i]))
}

length(which (MA.cv > Pop.cv)) # 37/39 kmers have lower CV
# sign test for significance
binom.test(37, 39)

par(mar=c(8,4,1,1))
# this is Figure 4C
plot(MA.cv, col="red", xlab="", ylab="CV", ylim=c(0,max(MA.cv)+10), type="l", xaxt="n")
lines(Pop.cv, col="blue")
axis(at=c(1:length(common.kmer.indexes)), side=1, labels=names(common.kmer.GC), las=2, cex.axis=0.4)
legend(x="top", c("MA lines", "Population"), lty=c(1,1), col=c("red","blue"), pt.cex=1, cex=0.7, bty="n")
```
### Quality control
Ensure the difference in CV is not because variance in increased in the MA lines because the second library is very different from the first library.   
Recalculate the CV of only library 1 - are most still more variable in the MA lines than the population?

```{r}
MA1.cv <- c()
for (i in common.kmer.indexes) {
  MA1.cv <- c(MA1.cv, cv(daphniakmers_normalized[MA1_indexes,i]))
}

Pop.cv <- c()
for (i in common.kmer.indexes){
  Pop.cv <- c(Pop.cv, cv(daphniakmers_normalized[Pop_indexes,i]))
}

pop_generations <- 62

corr_factor <- pop_generations/mean_generations
length(which((MA1.cv*corr_factor) > Pop.cv))
# sign test
binom.test(25, 39)

```
# Mutation rate of MA lines versus population  
## Calculate the mutation rate in the population and compare  
See supplementary information for how the conservative generation of 62 was estimated.  

``` {r}
# use 62 generations as estimate
# MA lines underwent 101 generations when population isolates collected, which was approximately one year after the MA lines were sequenced

pop_generations <- 62
mutation.rate.abs.pop.norm <- colMeans(kmer.deviation.abs[Pop.indexes,mutation.kmer.indexes])/(pop_generations*initial_means[mutation.kmer.indexes])

length(which(mutation.rate.abs.norm > mutation.rate.abs.pop.norm)) # 32/39 kmers have a lower realized mutation rate in the population than the MA lines
# sign test for significance
binom.test(32, 39)
```
  
### Plot the mutation rate in MA lines and population isolates
  
``` {r}
# This is Fig 4D
par(mar=c(8,4,1,1))
plot (mutation.rate.abs.norm, type="l", col="red", ylim=c(0,.008), xaxt="n", xlab="", ylab="Copies/generation/copy")
lines (mutation.rate.abs.pop.norm, col="blue")

axis(side=1, las=2, at=c(1:length(mutation.rate.abs.norm)), labels=names(mutation.rate.abs.norm), cex.axis=0.5)
legend(x="top", c("MA lines", "Population"), lty=c(1,1), col=c("red","blue"), pt.cex=1, cex=0.6, bty="n")

```
  
### Subsample the MA lines

```{r}
mutation.rate.pop.abs <- colMeans(kmer.deviation.abs[Pop.indexes,mutation.kmer.indexes])/(pop_generations)

mutation.rate.abs_mx <- kmer.deviation.abs[MA.indexes, mutation.kmer.indexes]/(generations)

Calc_CI_95 <- function (kmer_index) {
  means <- c()
  for (j in 1:1000) {
    k <- sample(MA.indexes, size=6, replace=T)
    means <- c(means, mean(mutation.rate.abs_mx[k, kmer_index]))
  }
  quantiles <- quantile(means, c(0.05, 0.95))
  return(quantiles)
}

lower_CIs_95 <- c()
for (i in 1:length(mutation.rate.pop.abs)) {
  lower_CIs_95 <- c(lower_CIs_95, Calc_CI_95(i)[1])
}

upper_CIs_95 <- c()
for (i in 1:length(mutation.rate.pop.abs)) {
  upper_CIs_95 <- c(upper_CIs_95, Calc_CI_95(i)[2])
}

which(mutation.rate.pop.abs < lower_CIs_95)
length(which(mutation.rate.pop.abs < lower_CIs_95))
# 22/38 kmers have a significantly lower mutation rate in the MA lines

```  
### Some kmers are more constrained than others  

``` {r}
mutation.rate.diffs <- (mutation.rate.abs.norm - mutation.rate.abs.pop.norm)/mutation.rate.abs.norm
mutation.rate.diffs[order(mutation.rate.diffs)]
# arbitrary measure for constraint to illustrate the variation in constraint level between different kmers.  
mutation.rate.diffs[which(mutation.rate.diffs < 0.10)] # which kmers have not a large constraint
mutation.rate.diffs[which(mutation.rate.diffs > 0.75)] # which kmers have a high constraint 

names(mutation.rate.diffs[which(mutation.rate.diffs > 0.75)])  %in% names(which(mutation.rate.pop.abs < lower_CIs_95)) 
# all of the highly constrained kmers are significant

```
# Kmers mutating in a correlated manner  

Make a correlation matrix of the changes in the kmer abundance.  
Compare between MA lines and population.
``` {r}
par(mar=c(8,4,1,1))
# Fig 5A
deviationCorr_MA <- cor(kmer.deviation[MA.indexes,order(common.kmer.GC)])
corrplot(deviationCorr_MA, type="upper", order="original", tl.col="black", tl.cex=0.4, tl.srt=45)

# Fig 5B
deviationCorr_pop <- cor(kmer.deviation[Pop.indexes,order(common.kmer.GC)])
corrplot(deviationCorr_pop, type="upper", order="original", tl.col="black", tl.cex=0.4, tl.srt=45)

# Fig 5C - check correlations between related kmers
related.kmer.indexes <- c(3,9,12,15,17,28,34,35,39)
temp <- kmer.deviation[MA.indexes,related.kmer.indexes]
colnames(temp) <- common.kmer.names[related.kmer.indexes]
deviationCorr_MA_rel <- cor(temp)
corrplot(deviationCorr_MA_rel, type="upper", order="original", tl.col="black", tl.cex=0.4, tl.srt=45)

```
  
## Mantel test shows difference between population and MA lines  

```{r}
mantel.test(deviationCorr_MA, deviationCorr_pop, graph = TRUE)
