---
title: "Quality control"
author: "Jullien Flynn"
date: '2017-01-20'
output:
  html_document:
    toc: yes
  pdf_document:
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Load necessary packages:
``` {r, load packages, include=FALSE}
## load packages
rm(list = ls()) # clear environment
library(devtools)
library(ggbiplot)
library(raster)
library(corrplot)
source("http://bioconductor.org/biocLite.R")
require(Biostrings)
library(adegenet)
library(ape)
library(Rcmdr)
library(MASS)
```
Quality control steps required:
* number of reads with kmers/total number of reads
* raw data before normalization , compare to normalized data
* compare technical replicates

# Reads with kmers per sample
``` {r}
kmer_reads <- read.csv("~/Documents/FirstRotation-Andy/kmer_reads.csv")
# last row is the total, get rid of it
kmer_reads <- kmer_reads[1:nrow(kmer_reads)-1,]
plot(x=kmer_reads$Total.reads, y=kmer_reads$kmer.reads, xlab="Total reads", ylab="Reads with kmers")
# top cluster is the second library prep, which has higher depth but the same proportion of reads with kmers as the first library.
```
# Compare before and after normalization  
Check correlation with kmer abundance to read depth.

``` {r}
# import raw, un-normalized data
daphniakmers <- read.delim("~/Documents/FirstRotation-Andy/daphniakmer.rep.compiled.txt")
# name the rows as the names of the lines
file_names <- as.vector(daphniakmers$lines)
file_names <- sapply(strsplit(file_names, "\\."), '[', 1) 
rownames(daphniakmers) <- file_names
# now get rid of the column that has the file name
daphniakmers[,1] <- NULL
# the columns (kmers) by their mean abundance
means <- colMeans(daphniakmers)
index.orders <- order(means, decreasing=T)
daphniakmers <- daphniakmers[,order(means, decreasing=T)]
# remove the total bp column and the poly-N columns
daphniakmers$total_bp <- NULL
daphniakmers$N.N <- NULL

# change the colnames to not show the reverse complement
kmer.labels <- sapply(strsplit(colnames(daphniakmers), "\\."), '[', 1)
colnames(daphniakmers) <- kmer.labels

# Get rowSums for each sample
raw_counts <- rowSums(daphniakmers)
raw_counts

# import read depth (calculated from GATK DepthOf Coverage)
mapdp <- read.table("~/Documents/mapdp.txt", header=F)
mapdp <- as.vector(mapdp[, 1])

# plot the raw total kmer counts against the average read depth
plot(x=mapdp, y=raw_counts, xlab="Average read depth", ylab="Total raw kmer counts")

# import data normalized by depth and GC-content
daphniakmers_normalized_w_techrep <- read.csv("~/Documents/FirstRotation-Andy/daphniakmers_corrected_bins3_Jan7.csv")
# process it
file_names <- as.vector(daphniakmers_normalized_w_techrep$X)
file_names <- sapply(strsplit(file_names, "\\."), '[', 1) 
rownames(daphniakmers_normalized_w_techrep) <- file_names
# now get rid of the column that has the file name
daphniakmers_normalized_w_techrep[,1] <- NULL
# the columns (kmers) by their mean abundance
means <- colMeans(daphniakmers_normalized_w_techrep)
index.orders <- order(means, decreasing=T)
daphniakmers_normalized_w_techrep <- daphniakmers_normalized_w_techrep[,order(means, decreasing=T)]
# remove the total bp column and the poly-N columns
daphniakmers_normalized_w_techrep$total_bp <- NULL
daphniakmers_normalized_w_techrep$N.N <- NULL

daphniakmers_normalized_w_techrep <- as.matrix(daphniakmers_normalized_w_techrep)
## need to convert to matrix
daphniakmers_normalized_w_techrep <- as.matrix(daphniakmers_normalized_w_techrep)
# don't deal with the technical replicates for now
techreps <- c("C001", "C035")
daphniakmers_normalized <- daphniakmers_normalized_w_techrep[!rownames(daphniakmers_normalized_w_techrep) %in% techreps, ]
total_normalized_counts <- rowSums(daphniakmers_normalized)

# plot this now
plot(x=mapdp, y=total_normalized_counts, xlab="Average read depth", ylab="Total normalized kmer counts")
```
There was a positive linear relationship between read depth and kmer counts before normalization, as to be expected.  
After normalizing, this positive relationship is no longer apparent. A slight negative relationship is likely because the population isolates had higher read depth, and they had less kmer content (which we argue is due to selection, see manuscript).

# Process this data without technical replicates and get summaries

``` {r}
common.kmer.indexes <- c()
for (i in 1:ncol(daphniakmers_normalized)) {
  if (mean(daphniakmers_normalized[,i])>=100) {
    common.kmer.indexes <- c(common.kmer.indexes, i)
  }
}
names(common.kmer.indexes) <- colnames(daphniakmers_normalized)[common.kmer.indexes]
common.kmer.names <- colnames(daphniakmers_normalized)[common.kmer.indexes]

length(common.kmer.indexes)

Calc_GC_content <- function (kmer_string) {
  string <- BString(kmer_string)
  gc.content <- c(letterFrequency(string, letters=c("CG"), as.prob=F)/(length(string)))
  return(gc.content)
}

common.kmer.gc <- sapply(common.kmer.names, Calc_GC_content)
length(common.kmer.gc)
# get rid of the dot before the GC in the name
names(common.kmer.gc) <- common.kmer.names
```

# Compare technical replicates

``` {r}

C35.index <- which(rownames(daphniakmers_normalized_w_techrep)=="C35")
C01.index <- which(rownames(daphniakmers_normalized_w_techrep)=="C01")
C001.index <- which(rownames(daphniakmers_normalized_w_techrep)=="C001")
C035.index <- which(rownames(daphniakmers_normalized_w_techrep)=="C035")

C01.comparison_common.kmers <- as.vector(daphniakmers_normalized_w_techrep[C01.index, common.kmer.indexes]) / as.vector(daphniakmers_normalized_w_techrep[C001.index, common.kmer.indexes])

names(C01.comparison_common.kmers) <- colnames(daphniakmers_normalized_w_techrep)[common.kmer.indexes]

C01.comparison_common.kmers
mean(C01.comparison_common.kmers)
sd(C01.comparison_common.kmers)

C35.comparison_common.kmers <- as.vector(daphniakmers_normalized_w_techrep[C35.index, common.kmer.indexes]) / as.vector(daphniakmers_normalized_w_techrep[C035.index, common.kmer.indexes])

names(C35.comparison_common.kmers) <- colnames(daphniakmers_normalized_w_techrep)[common.kmer.indexes]

C35.comparison_common.kmers
mean(C35.comparison_common.kmers)
sd(C35.comparison_common.kmers)

```
See if deviations are consistently related to GC content

```{r}
plot(C01.comparison_common.kmers, pch=19, col="purple", ylim=c(0.2, 2) , ylab="ratio")
points(C35.comparison_common.kmers, pch=19, col="pink")


plot(y=C01.comparison_common.kmers, x=common.kmer.gc, type="p", col="purple", pch=19, ylim=c(0.2, 2.0), xlab="GC content", ylab="ratio")
points(y=C35.comparison_common.kmers,x=common.kmer.gc, pch=19, col="pink" )

```
Conclusion: most of the ratios are consistent and close to 1, however there is unpredictable variation for kmers at 60% GC.

## Compare correlation in technical replicates to correlation among MA lines
?? Problem: not more correlated with each other than the 
``` {r}
(t(daphniakmers_normalized_w_techrep[, common.kmer.indexes]))
cor_MAlines <- cor(t(daphniakmers_normalized_w_techrep[, common.kmer.indexes]))
cor_MAlines
# all MA lines are highly correlated
cov_MAlines <- cov(t(daphniakmers_normalized_w_techrep[, common.kmer.indexes]))
cov_MAlines
```

## Do a PCA of the 3 library preps
``` {r}
lib1.lines <- c("C02", "C03", "C06", "C07", "C08", "C12", "C13", "C14", "C16", "C17", "C18", "C20", "C21", "C24", "C25", "C34", "C36", "C37", "C38", "C39", "C40", "C44")
lib1.indexes <- match(lib1.lines, rownames(daphniakmers_normalized_w_techrep))
lib2.lines <- c("C23", "C27", "C43", "C49", "CC3", "CC4", "CC6", "CC7", "CC8", "CC9")
lib2.indexes <- match(lib2.lines, rownames(daphniakmers_normalized_w_techrep))
C01.index
C001.index
C35.index
C035.index
lib1.indexes
lib2.indexes

pca.data <- prcomp(daphniakmers_normalized_w_techrep[, common.kmer.indexes], scale.=T)
category <- factor(c( c("C001", "C01", "lib1", "lib1", "C035"), c(rep("lib1",times=11)), "lib2", "lib1", "lib1", "lib2", "lib1", "C35", c(rep("lib1", times=5)), "lib2", "lib1", c(rep("lib2", times=7)) ))

g <- ggbiplot(pca.data, obs.scale = 1, var.scale = 1, groups = category, var.axes = F)
g <- g + scale_color_discrete(name='')
g <- g + theme(legend.direction='horizontal', legend.position='top')
print(g)
```
Conclusion: PC1 is separating samples based on a biological factor (mutations). PC2 is separating samples based on library prep. Technical replicates are very close along PC1 and a bit separated along PC2.

## Use permutations to see if there is a significant difference in the abundance of each kmer for the different libraries
``` {r}
MA1_names <- c("C01", "C02", "C03", "C06", "C07", "C08", "C12", "C13", "C14", "C16", "C17", "C18", "C20", "C21", "C24", "C25", "C34", "C35", "C36", "C37", "C38", "C39", "C40", "C44")
MA1_indexes <- match(MA1_names, rownames(daphniakmers_normalized))
MA2_names <- c("C23", "C27", "C43", "C49")
MA2_indexes <- match(MA2_names, rownames(daphniakmers_normalized))

MA1.means <- c()
for (i in common.kmer.indexes) {
  MA1.means <- c(MA1.means, mean(daphniakmers_normalized[MA1_indexes,i]))
}
MA2.means <- c()
for (i in common.kmer.indexes) {
  MA2.means <- c(MA2.means, mean(daphniakmers_normalized[MA2_indexes,i]))
}

# calculate differences in the means of MA1 and MA2
means_diffs <- (MA1.means - MA2.means)/MA1.means
plot(means_diffs)
abline(h=0, col="red")


## randomly sample 4 of the 28 and calculate the mean. Do this 1000 times for each kmer. 
# calculate the 0.95 and 0.05 quantile of the distribution.
#If the 95% CI is wider than the difference between them, then good.
Calc_CI <- function (kmer_index) {
  means <- c()
  for (j in 1:1000) {
    k <- sample(c(1:24), size=4, replace=T)
    means <- c(means, daphniakmers_normalized[k, kmer_index])
  }
  quantiles <- quantile(means, c(0.05, 0.95))
  return(quantiles)
}


lower_CIs <- c()
for (i in common.kmer.indexes) {
  lower_CIs <- c(lower_CIs, Calc_CI(i)[1])
}

upper_CIs <- c()
for (i in common.kmer.indexes) {
  upper_CIs <- c(upper_CIs, Calc_CI(i)[2])
}

plot(MA1.means[1:10], type="l", col="red", ylim=c( min(lower_CIs[1:10]), max(upper_CIs[1:10])))
lines(MA2.means[1:10], col="blue")
lines (lower_CIs[1:10], lty=2)
lines (upper_CIs[1:10], lty=2)

plot(MA1.means[11:23], type="l", col="red", ylim=c( min(lower_CIs[11:23]), max(upper_CIs[11:23])))
lines(MA2.means[11:23], col="blue")
lines (lower_CIs[11:23], lty=2)
lines (upper_CIs[11:23], lty=2)

plot(MA1.means[24:length(common.kmer.indexes)], type="l", col="red", ylim=c( min(lower_CIs[24:length(common.kmer.indexes)]), max(upper_CIs[24:length(common.kmer.indexes)])))
lines(MA2.means[24:length(common.kmer.indexes)], col="blue")
lines (lower_CIs[24:length(common.kmer.indexes)], lty=2)
lines (upper_CIs[24:length(common.kmer.indexes)], lty=2)

v <- c()
for (i in common.kmer.indexes) {
  v <-  c(v, (min(MA1.means[i], MA2.means[i]) > (Calc_CI(i)[1])) & (max(MA1.means[i], MA2.means[i]) < (Calc_CI(i)[2])))
}
sum(v==TRUE)

```

## Is the variation between technical replicates less than the variation between samples?
``` {r}
# try a paired wilcoxon signed test

C01.C001.df <- data.frame(daphniakmers_normalized_w_techrep[C01.index, common.kmer.indexes], daphniakmers_normalized_w_techrep[C001.index, common.kmer.indexes])
colnames(C01.C001.df) <- c("C01", "C001")

wilcox.test(C01.C001.df$C01, C01.C001.df$C001, paired = TRUE)


C35.C035.df <- data.frame(daphniakmers_normalized_w_techrep[C35.index, common.kmer.indexes], daphniakmers_normalized_w_techrep[C035.index, common.kmer.indexes])
colnames(C35.C035.df) <- c("C35", "C035")

wilcox.test(C35.C035.df$C35, C35.C035.df$C035, paired = TRUE)

# not significant difference between technical replicates

# significant difference between these two MA lines in the same technical replicate.
# do pairwise comparisons of MA lines??

C01.C35.df <- data.frame(daphniakmers_normalized_w_techrep[C01.index, common.kmer.indexes], daphniakmers_normalized_w_techrep[C35.index, common.kmer.indexes])
colnames(C01.C35.df) <- c("C01", "C35")

result <- wilcox.test(C01.C35.df$C01, C01.C35.df$C35, paired = TRUE)
result$p.value

C001.C035.df <- data.frame(daphniakmers_normalized_w_techrep[C001.index, common.kmer.indexes], daphniakmers_normalized_w_techrep[C035.index, common.kmer.indexes])
colnames(C001.C035.df) <- c("C001", "C035")

result <- wilcox.test(C001.C035.df$C001, C001.C035.df$C035, paired = TRUE)
result$p.value
result

wilcoxp_matrix <- matrix(ncol=nrow(daphniakmers_normalized_w_techrep), nrow=nrow(daphniakmers_normalized_w_techrep))

for (i in 1:nrow(daphniakmers_normalized_w_techrep)) {
  for (j in 1:nrow(daphniakmers_normalized_w_techrep)) {
    df <- data.frame(daphniakmers_normalized_w_techrep[i, common.kmer.indexes], daphniakmers_normalized_w_techrep[j, common.kmer.indexes])
    result <- wilcox.test(df[,1], df[,2], paired = TRUE)
    wilcoxp_matrix[i,j] <- result$p.value
  }
}

hist(wilcoxp_matrix)

alpha <- 0.05/(nrow(wilcoxp_matrix) -1)
alpha
length(which(wilcoxp_matrix < 0.05))


